A Theory for Migration of Information Systems

### 2022-08-28 2022 Migration/articleMigration.tex
Processing the changes discussed on 2022/08/28

\documentclass{elsarticle}
\usepackage{graphicx}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\include{preamramics}
\hyphenation{ExecEngine}
\newtheorem{lemma}{Lemma}
\def\id#1{\text{\it #1\/}}
\def\Events{{\mathit E}}
\begin{document}

@@

\section{Basics}
\label{sct:Basics}
	In this section we explain the basic notions to describe migration.
	We introduce the notion of dataset.
	A dataset is a quintuple $\la\rules,\rels,\concepts,\pairs,\atoms\ra$,
	which consists of a set of rules $\rules$, a set of relations $\rels$, a set of concepts $\concepts$, a set of pairs $\pairs$, and a set of atoms $\atoms$.
	Let us introduce these five notions one by one.

	Atoms are values that have no internal structure, meant to represent data elements in a database.
	From a business perspective, atoms are used to represent concrete items of the world,
	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
	The set of all atoms is called $\atoms$.
    Each atom is an instance of a \emph{concept}.

	Concepts (from set $\concepts$) are names we use to classify atoms in a meaningful way.
	For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
	In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
	The expression $a \in A$ means that atom $a$ is an \emph{instance} of concept $A$.
	The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
	Specialization is needed to allow statements such as: ``A woman is a human.'' or ``A human is a mammal.''.

	Pairs (from set $\pairs$) are used to relate two atoms.
	A pair $\pair{\atom{Peter}}{\atom{074238991}}$ might bear the meaning that the ``thing'' that \atom{Peter} refers to
	has \atom{074238991} as a telephone number.
	In the formal world, we do not care about that meaning,
	leaving it entirely up to a user which meaning (s)he attaches to this pair.

	Relations (from set $\rels$) are used in information systems to store data as pairs.
	In this paper relations are represented by variables $r$, $s$, and $d$.
	Every relation has a source concept and a target concept.
	If we need to be specific about the source and target concepts, we write $\declare{r}{A}{B}$ instead of just $r$.
	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concepts $A$ and $B$.
	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.
	The meaning of relations is defined by an interpretation function $\mathfrak{I}$.
	It maps each relation to a set of pairs.

	The typing rule guarantees that each pair in $r$ respects the type of $r$:
\begin{equation}
	\tag{typing rule}
	\pair{a}{b}\in\ti{\declare{r}{A}{B}} \Rightarrow\ a \in A \wedge b \in B \label{typing of relations}
\end{equation}

\begin{equation}
	\tag{population}
	\pairs\ =\ \{ \pair{a}{b}|\ r\in\rels,\ \pair{a}{b}\in\ti{r}\} \label{def. population}
\end{equation}

\begin{equation}
	\tag{atoms}
	\atoms\ =\ \{ a |\ \pair{b}{c}\in\pairs,\ a\in\{b,c\}\ \} \label{def atoms}
\end{equation}

	A rule is an equality between two terms that constrains the data stored in relations,
	to ensure that the dataset is at all times consistent with all rules in the dataset ($\rules$).
	Terms are used to combine relations using operators.
	The set of terms is called $\terms$.
\begin{definition}[terms]
\label{def:terms}
\item   The set of terms, $\terms$, is the smallest set that satisfies, for all $r,s \in \terms$, $d\in\rels$ and $A,B \in \concepts$: 
\begin{eqnarray}
	d&\in&\terms         \quad\quad\text{(every relation is a term)}\\
	(r \cup s)&\in&\terms\quad\quad\text{(union)}\\
	(r \cap s)&\in&\terms\quad\quad\text{(intersection)}\\
	(r-s)&\in&\terms     \quad\quad\text{(difference)}\label{def:difference}\\
	(r;s)&\in&\terms     \quad\quad\text{(composition)}\\
	(r\backslash s)&\in&\terms     \quad\quad\text{(right residual)}\\
	(r\slash s)&\in&\terms     \quad\quad\text{(left residual)}\\
	\flip r&\in&\terms   \quad\quad\text{(converse)}\\
	\cmpl{r}&\in&\terms   \quad\quad\text{(complement)}\\
%	\kleeneplus r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
%	\kleenestar r&\in&\terms   \quad\quad\text{(Kleene closure)}\\
	\ident{A}&\in&\terms \quad\quad\text{(identity)}\\
	\fullt{A}{B}&\in&\terms \quad\quad\text{(full set)}
\end{eqnarray}
\end{definition}
	Throughout the remainder of this paper,	terms are represented by variables $r$, $s$, $d$, and $t$.
	The \define{type} of a term $r$ is a pair of concepts, given by $\tf{r}$.
	$\mathfrak{T}$ is a partial function that maps terms to types.
	The type function and its restrictions are discussed in~\cite{Joosten2015}
	and have no consequences for the remainder of this paper.
	(\emph{\@Bas, komen we ermee weg als we de definitie van $\tf{r}$ hier weglaten?})

 	The meaning of terms is given by an interpretation function $\mathfrak{I}$.
	Let $A$ and $B$ be finite sets of atoms, then $\mathfrak{I}$ maps each term to the set of pairs for which that term stands.
\begin{definition}[interpretation of terms]
\label{interpretation of terms}
\item   For every $A,B\in\concepts$ and $r,s\in\terms$
\begin{eqnarray}
	\ti{r}		 &=&\{\pair{a}{b}|\ a\ r\ b\}	\\
	\ti{r \cup s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{or }\ \pair{a}{b}\in\ti{s}\}	\\
	\ti{r \cap s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\in\ti{s}\}	\\
	\ti{r-s}	 &=&\{\pair{a}{b}|\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{a}{b}\notin\ti{s}\}	\\
	\ti{r;s}	 &=&\{\pair{a}{c}|\ \text{for some}\ b,\ \pair{a}{b}\in\ti{r}\ \text{and}\ \pair{b}{c}\in\ti{s}\}	\\
	\ti{r\backslash s}	 &=&\{\pair{b}{c}|\ \text{for all}\ a,\ \pair{a}{b}\in\ti{r}\ \text{implies}\ \pair{a}{c}\in\ti{s}\}	\\
	\ti{r\slash s}	 &=&\{\pair{a}{b}|\ \text{for all}\ c,\ \pair{b}{c}\in\ti{s}\ \text{implies}\ \pair{a}{c}\in\ti{r}\}	\\
	\ti{\flip{r}}	 &=&\{\pair{b}{a}|\ \pair{a}{b}\in\ti{r}    \}	\\
	\ti{\cmpl{\declare{r}{A}{B}}}	 &=&\fullt{A}{B}-r	\\
%	\ti{\kleeneplus{r}}	 &=&\text{the smallest set that satisfies }\kleeneplus{r}=r\ \cup\ \kleeneplus{r};\kleeneplus{r}	\\
%	\ti{\kleenestar{\declare{r}{A}{A}}}	 &=&\ident{A}\cup\kleeneplus{r}\\
	\ti{\ident{A}} 	 &=&\{\pair{a}{a}|\ a\in A\}	\\
	\ti{\fullt{A}{B}}&=&\{\pair{a}{b}|\ a\in A, b\in B\}
\end{eqnarray}
\end{definition}
%	The Kleene closure operators (postfix $\kleeneplus{\ }$ and $\kleenestar{\ }$) have been implemented partially in the current Ampersand implementation.

	A \define{rule} is a pair of terms $r,s\in\terms$ with the same types ($\tf{r}=\tf{s}$).
\[\text{RULE}\ r = s\]
	This means \(\ti{r} = \ti{s}\). In practice, many rules are written as:
\[\text{RULE}\ r\subs s\]
	This is a shorthand for 
\[\text{RULE}\ r\cap s = r\]
	We have enhanced the type function $\mathfrak{T}$ to cover rules as well.
	If $\tf{r}=\tf{s}$ and $\tf{s}=\pair{A}{B}$:
\begin{eqnarray}
	\tf{\text{RULE}\ r = s}   &=&\pair{A}{B}\\
	\tf{\text{RULE}\ r\subs s}&=&\pair{A}{B}
 \end{eqnarray}

	We call a rule $\text{RULE}\ r = s$ \define{satisfied} if \(\ti{r} = \ti{s}\).

	The data in a data set $\la\rules,\rels,\concepts,\pairs,\atoms\ra$ are the pairs in the interpretation of all relations in $\rels$.
	This constitutes the set $\pairs$, also known as the population of the dataset.
	In response to messages such as
\[\text{INSERT}\ <3,{\tt Brown}> \text{INTO}\ \declare{r}{Age}{Name}\]
 	the population of a dataset changes. 
	As the population of relations used in $r$ changes with time, the satisfaction of the rule changes accordingly.
	A software developer, who conceives these rules, must consider how to keep each one of them satisfied.
	We call a rule \define{violated} if it is not satisfied.
	The set $\ti{(s-r)\cup(r-s)}$ is called the \emph{violation set} of \(\text{RULE}\ r = s\).
	To \define{resolve} violations means to change the contents of relations such that the rule is satisfied%
\footnote{To \define{restore invariance} is sometimes used as a synonym to resolving violations.
	To \define{preserve invariance} is sometimes used as a synonym to keeping a rule satisfied.}.
	Consequently, such a rule may also be referred to as \define{invariant}.
	Each pair in the violation set of a rule is called a violation of that rule.

>>>>

\documentclass{elsarticle}
\usepackage{graphicx}
%\usepackage{multicol}
%\usepackage{footmisc}
\usepackage{amstext}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[english]{babel}
%\usepackage[official,right]{eurosym}
\selectlanguage{english}
\hyphenation{ExecEngine}
\newtheorem{lemma}{Lemma}
\def\Events{{\mathit E}}
\begin{document}
\include{preambleMigrations}

@@

\section{Basics}
\label{sct:Basics}
	To describe migration, we first introduce the notion of dataset.
	A dataset describes a set of data, which is typically stored persistently in a database of some kind.
	Before defining datasets, we must first define the constituent notions of atom, concept, relation, rule, and role.
\subsection{Atoms}
	Atoms are values that have no internal structure, meant to represent data elements in a database.
	From a business perspective, atoms are used to represent concrete items of the world,
	such as \atom{Peter}, \atom{1}, or \atom{the king of France}.
	By convention throughout the remainder of this paper, variables $a$, $b$, and $c$ are used to represent \emph{atoms}.
	The set of all atoms is called $\atoms$.

\subsection{Concepts}
	Concepts are names we use to classify atoms in a meaningful way.
	For example, you might choose to classify \atom{Peter} as a \concept{Person}, and \atom{074238991} as a \concept{TelephoneNumber}.
	In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
    We will use variables $A$, $B$, $C$, $D$ to represent concepts.
	The term $\ident{A}$ represents the \emph{identity relation} of concept $A$.
	The expression $a \in A$ means that atom $a$ is an \emph{instance} of concept $A$.
	The statement $A\isa B$ (pronounce: $A$ is a $B$) states that any instance of $A$ is an instance of $B$ as well.
	We call this {\em specialization}, but it is also known as {\em generalization} or {\em subtyping}.
	Specialization is needed to allow statements such as: ``A woman is a human'' or ``A human is a mammal''.
	The set of all concepts is called $\concepts$.

\subsection{Triples}
	A triple $\triple{\text{\atom{Peter}}}{\id{phone}}{\text{\atom{074238991}}}$ might mean that the ``thing'' that \atom{Peter} refers to
	has \atom{074238991} as a telephone number.
	Triples are used to relate two atoms and a relation.
	In the formal world, we leave it entirely up to a user to attach meaning to a triple.
	The practical meaning has no consequences in the formal world.
	The set of all triples is called $\triples$.

\subsection{Relations}
	Relations are used in information systems to store data.
	In this paper relations are represented by variables $r$, $s$, and $d$.

	Every relation $r$ has a set of pairs, which we call the population of $r$:
\begin{equation}
	\pairs{r}\ =\ \{ \pair{a}{b}|\ \triple{a}{r}{b}\in\popR\}
\end{equation}
	Every relation has a name, a source concept, and a target concept.
	We write $\declare{nm}{A}{B}$ to denote a relation with name \id{nm}, source concept $A$, and target concept $B$.
	% To disassemble a relation in its name, source and target, there exist three functions:
% \[\begin{array}[3]{rcl}
	% \id{relname}&:&\rels\rightarrow\id{RelationIdentifier}\\
	% \id{src}&:&\rels\rightarrow\concepts\\
	% \id{tgt}&:&\rels\rightarrow\concepts
% \end{array}\]
	% Or, stated otherwise, the denotation $\declare{nm}{A}{B}$ stands for a relation with
	% \(\relname{\declare{nm}{A}{B}} = \id{nm}\),
	% \(\src{\declare{nm}{A}{B}}=A\), and
	% \(\tgt{\declare{nm}{A}{B}}=B\).
	The pair $\pair{A}{B}$ is called the \emph{signature} of the relation.
	The term $\fullt{A}{B}$ represents the \emph{universal relation} over concept symbols $A$ and $B$.
	The term $\ident{A}$ represents the \emph{identity relation} over concept $A$.

\subsection{Roles}
	A role is a name that identifies a group of users that get the same privileges in a data set.
	The set of all roles is called $\roles$.

\subsection{Constraints}
	Each constraint $u\in\rules$ has a signature $\signat{u}$, which is a pair of concepts.
	Each constraint $u$ also has a function $\id{viol}_u:\powerset{\atoms\times\rels\times\atoms}\ \rightarrow\ \powerset{\atoms\times\atoms}$.
	We call the set $\viol{\popR}{u}$ the \define{violation set} of constraint $u$ in population $\popR$.
	We call every element $\pair{a}{b}\in\viol{\popR}{u}$ a \define{violation} of constraint $u$ in $\popR$.
	Note that the signature of $u$ is the same for every population, whereas the violation set may be different for every $\popR$.

\subsection{Datasets}
	We can now define the notion of dataset.
	Let
	$\concepts$ be a set of concept symbols,
	$\rels$ be a set of relation symbols,
	$\roles$ be a set of roles,
	$\rules$ be a set of constraints, and
	$\atoms$ be a set of atoms.
\begin{definition}[dataset]
	\label{def:dataset}
	\item A dataset is a tuple $\la\concepts,\rels,\roles,\rules,\maintain,\atoms,\popR,\popC\ra$ with
	\[\begin{array}[3]{rcl}
		\id{maintain}&:&\roles\times\rules\\
		\id{popR}&:&\atoms\times\rels\times\atoms\\
		\id{popC}&:&\concepts\times\atoms\\
		\multicolumn{3}{l}{\triple{a}{r}{b}\in\popR\ \Rightarrow\ \pair{\src{r}}{a}\in\popC \wedge \pair{\tgt{r}}{b}\in\popC}\\
		\forall a\in\atoms\ \exists A\in\concepts&:&\pair{A}{a}\in\popC
	\end{array}\]
\end{definition}
